---
title: "Creating Maps in R"
output: html_document
---

This demo will use crime data from opendata philly recorded between 2012 and 2014. The first step is to import the libraries I will be using to process the data. I will import the first 50,000 rows of the Philadelphia Crime Data. Note: If you want to import the complete file you would just delete the <i>n_max</i> statement.  

```{r, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(lubridate)
library(RColorBrewer)
library(leaflet)
library(stringr)
library(rgdal)
library(readr)
source('fte_theme.R')
source('multiplot.R')

# ds <- read_csv('https://data.phila.gov/api/views/7ret-xhtc/rows.csv?accessType=DOWNLOAD', n_max = 50000)
# local copy import
ds  <- read_csv('data/PPD_Crime_Incidents_2012-2014.csv', n_max = 50000)
```

The <i>read_csv</i> function stores the data in a <i>dataframe</i>.  A dataframe is basically a table which can store different types of data. The first 5 rows of the the dataframe <i>ds</i> are shown below using the <i>head</i> function. 
```{r}
head(ds)
```

So the first issue with this dataset is the way the coordinates are presented. They are provided as a string surrounded by parentheses. 
```{r}
head(ds$Coordinates)
```

To deal with this I will have to use some regular expressions commands.  Regular expression commands allow you to parse text data in a variety of ways. If you don't have much experience with regular expression commands check out [Sam Hughes post on the subject](http://qntm.org/files/re/re.html) and [my gist](https://gist.github.com/ctufts/14e75172303dec0218ba) with some examples you can run in R. 

```{r}
# delete opening parenthesis - sub ( with nothing
ds$Coordinates<- gsub(")", "", ds$Coordinates)
# delete closing parenthesis - sub ) with nothing
ds$Coordinates<- gsub("\\(", "", ds$Coordinates)
# split each coordinate entry at the comman
coord.list <- (str_split(ds$Coordinates, ","))
# first entry to the x coordinate
y <- as.numeric(unlist(lapply(coord.list, function(x)x[1])))
# second entry to the y coordinate
x <- as.numeric(unlist(lapply(coord.list, function(x)x[2])))
# create new data frame with the new coordinates
# the crime type, district, and the date/time of the crime
coordinates <- data.frame(POINT_Y = y,
                          POINT_X = x,
                          General.Crime.Category = ds$`General Crime Category`,
                          District = ds$District,
                          Dispatch.Date = ds$`Dispatch Date/Time`
)
```
I am removing any rows The 22nd and 23rd districts were combined years back, so I am going to combine the two. 

```{r}
# remove any rows with missing coordinates
coordinates <- na.omit(coordinates)
# combine 22nd and 23rd districts
coordinates$District[coordinates$District==23] <- 22
```

The first map will display the number of crime incidents in each police district. To start this procedure I will have to calculate the number of crimes in each district.  

```{r}
# read in shape data for the police districts
districts <- readOGR(dsn = "data/shp",
                     layer = "Police_Districts", verbose = FALSE)

# count the number of crime committed by each district
# group by crime type
district.log <- coordinates %>% group_by(General.Crime.Category, District) %>%
  summarise(
    events = n()
  )
```

I created a dataframe named <i>district.log</i> which contains the number of crime incedents in each district grouped by crime type.  I will only focus on <i>Arson</i> events for the visualization. 
```{r}
##### select a type of crime #####
# select the arson crimes in the list
crime.type.selected <- 'Arson'
print(crime.type.selected)
# count the number of crimes committed in each district
district.counts <- filter(district.log, General.Crime.Category == crime.type.selected) %>% 
  arrange(District)
```

Ok, so now I'm ready to plot a map. I create a color palette for the shading of the districts and then create the map. The color of each district is based on the number of arsons in the district.

```{r}

# create a color palette for the map
pal <- colorNumeric(
  palette = "Greys",
  domain = district.counts$events
)

# create map
# add a template set of tiles
# set the initial view positions and zoom level
# add polygons
crime.map <- leaflet() %>%
  addProviderTiles('CartoDB.Positron') %>%
  setView(lng=-75.16048, lat=39.99000, zoom =11) %>%
  addPolygons(data = districts,
              stroke = T, smoothFactor = 0.2, fillOpacity = 0.5,
              color = "#000000", weight = 2, 
              fillColor = ~pal(district.counts$events)
              # color = ~pal(states@data$DISTRICT_)
  )
crime.map
```

To get a better understanding of the number of arsons in each district; I add a legend.  

```{r}
crime.map <- crime.map %>% 
  addLegend("bottomright", pal = pal, values = district.counts$events,
            title = "Incidents 2012-2014",
            opacity = 1
  )
crime.map
```

Now I will add individual markers where each arson occurred. 

```{r}
point.coord <- filter(coordinates, General.Crime.Category == crime.type.selected)
addCircles(crime.map, lat = point.coord$POINT_Y, lng = point.coord$POINT_X)    

```

I want to change the color and radius of the markers. 

```{r}
addCircles(crime.map, lat = point.coord$POINT_Y, lng = point.coord$POINT_X,
             fillOpacity = 0.8 , opacity = 0.8, radius = 160, fillColor = "#F0F340",
             weight = 2,
             color = "#5E610B")    

```

The final touch is to add pop-up labels to the markers. 

```{r}
addCircles(crime.map, lat = point.coord$POINT_Y, lng = point.coord$POINT_X,
             fillOpacity = 0.8 , opacity = 0.8, radius = 160, fillColor = "#F0F340",
             color = "#5E610B", weight = 2,
             popup = paste("District:",point.coord$District, "<br/> Time:", point.coord$Dispatch.Date))    

```